\name{}: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-03-28 paper/main.tex
.

\documentclass{bioinfo}
\usepackage[english]{babel}
\copyrightyear{2015} \pubyear{2015}
\usepackage{natbib}
\bibliographystyle{natbib.bst}

@@

\subtitle{Subject Section}
\newcommand{\name}{Voodoo }

\title[short Title]{\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction}
\author[Sample \textit{et~al}.]{Tilman Hinnerichs\,$^{\text{\sfb 1,}*}$ and Robert Hoehndorf\,$^{\text{\sfb 2}}$}
\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
$^{\text{\sf 2}}$Department, Institution, City, Post Code,
Country.}

@@

\section{Methods}
\subsection{Problem Description}
\name aims to solve the following problem: for a given drug and a
given protein we want to determine whether those interact or not.  We
do not differentiate between types of interaction such as activation
and inhibition, and do not predict the strength of the interaction.
If we additionally assume that our knowledge is complete and all
drug--protein pairs without a known interaction do not interact, we
can formulate the problem as a binary classification task.

@@

\subsection{\name: computational model to identify drugs that target a
  protein}

@@

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank drugs that likely target this protein. \name combines two types
of features: structural information for drugs and proteins that can be
used to determine if the drug and protein physically interact, and
information about phenotypic effects of drugs and changes in protein
function that may ``localize'' on an interaction network (i.e.,
neighboring nodes will share some of these features or are
phenotypically similar).  As structural features, \name uses
structural representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020} and
ontology-based annotations of drugs and proteins.

We construct a graph with proteins as nodes and protein-protein
interactions as edges, mapping the protein features to each target as
node features. \name then propagates information among the PPI network
utilizing graph convolutional steps, calculates the similarity of drug
and protein representations, and predicts whether there is an
interaction. The full workflow scheme is depicted in
figure~\ref{fig:ModelWorkflow}

@@

\begin{table*}[!tpb]
  \centering
	\begin{tabular}{|p{2.2cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|}
		\hline
		&\multicolumn{4}{c|}{(a) STITCH results}& \multicolumn{4}{c|}{(b) Yamanishi results}\\
		\hline
		\name results&\multicolumn{4}{c|}{PPI graph}&\multicolumn{4}{c|}{PPI graph}\\
		&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
		&Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		MolPred&$0.69$&$0.65$&$0.69$&$0.67$&$0.66$&$0.67$&$0.66$&$0.64$\\
		\hline
		OntoPred&$0.88$&$0.87$&$0.92$&$0.93$&$0.80$&$0.79$&$0.83$&$0.82$\\
		\hline
		\name (MolPred + OntoPred) & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$& $0.83$ & $0.82$&$\mathbf{0.84}$&$\mathbf{0.84}$\\
		\hline
	\end{tabular} \medskip

	\begin{tabular}{|l|p{1cm}|p{1cm}|p{1cm}|p{1cm}|}
		\hline
		(c) Approach&\multicolumn{2}{c|}{Original scheme}&\multicolumn{2}{c|}{Protein split}\\
		&Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		Naive predictor&DP pairs&$0.85$& --&--\\
		DTINet&DP pairs&$0.91$&$0.74$&$0.67$\\
		DTIGEMS+&DP pairs&$\mathbf{0.93}$& $0.72$& $0.68$ \\
		DTI-CDF&Proteins&$0.85$&$\mathbf{0.85}$&$0.79$\\
		\name&Proteins&$0.84$&$0.84$&$\mathbf{0.84}$\\
		\hline
		\hline
		(d) Approach&\multicolumn{2}{c|}{Original scheme}&\multicolumn{2}{c|}{Protein split}\\
		&Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		Naive predictor&DP pairs&$0.79$&-- &--\\
		DeepDTI&Drugs&$0.88$&$0.76$&$0.70$\\
		DeepDTA&DP pairs&$0.88$&$0.77$&$0.69$\\
		DeepConv-DTI&DP pairs&$0.88$&$0.76$&$0.73$\\
		MolTrans&DP pairs&$0.90$&$0.77$&$0.74$\\
		\hline
	\end{tabular}\\

	\caption{(a) + (b) Results for \name on STITCH and Yamanishi dataset evaluated with a 5-fold cross-validation. We hereby denote the molecular feature based predictor with \textit{MolPred}, while abbreviating the ontology based, top-down predictor with \textit{OntoPred}. (c) + (d) Results for various state of the art (c) drug--target interaction prediction methods on Yamanishi dataset and (d) drug--target affinity prediction methods on BIOSNAP dataset, evaluated on their original and the protein cross-validation splitting scheme, approximately reproducing the results of MolTrans \textbf{CITATION}.}
	\label{tab:Results}
\end{table*}

@@

\subsection{Protein-centric evaluation}
The goal of \name is to find candidate drugs that target a specific
protein; however, so far, we do not evaluate this application but
rather how \name would perform in finding plausible drug--target
interactions among all possible interactions (since we use the
MacroAUC as our main evaluation measure). This evaluation does not
correspond to the application of \name in finding drugs that target a
specific protein.  To provide a better estimate on how \name performs
for individual protein targets, we use micro-averages between proteins
and compute the \textit{MicroAUC} (see Section \ref{sec:evaluation});
to determine {\em MicroAUC}, we average the performance (true and
false positive rates) per protein instead of across all drug--protein
pairs; the resulting measure can therefore better estimate how \name
performs when tasked with finding a drug that targets a specific
protein.

@@

The way in which we formulated the na\"ive predictor, it is not
applicable for a protein split cross-validation as the number of
interactions for each protein in the validation set is unknown.  We
apply this naive predictor on both the STITCH and Yamanishi datasets,
using the full datasets as well as a 5-fold cross-validation over
drugs and over drug--protein pairs to compare the prediction results
directly to \name.  For each fold, we gradually increase $k$ to
determine the best performance for each fold. Using the full dataset,
drug--target split, and drug split, we obtain the following MacroAUC
results\todo[inline]{Put in Results table, then remove from here}: for
the STITCH database, we obtain a performance of $0.76$ on the whole
dataset, $0.70$ for the drug--target pairs and $0.73$ in case of the
drug splitting scheme; on the Yamanishi dataset, we obtain MacroAUC
scores of $0.88$, $0.84$ and $0.85$, for the total dataset,
drug--target pair and drug split, respectively.  The na\"ive predictor
shows higher performance on the Yamanishi dataset than on STITCH, and
a substantial gain in comparison to an expected random predictor on
both datasets. In the following, we utilize this nai\"ve predictor as
baseline to compare its performance to state of the art models and
\name.

@@

The results of our experiments are summarized in part (c) of Table
\ref{tab:Results}\todo{Check after restructure}; we calculated the
performance of all compared methods over their original splitting
scheme and over a protein split. We find that there is a large
difference in performance when evaluating over a drug--target pair
split compared to a protein split, with generally higher performance
achieved when using the drug--target pair split.  Second, when
evaluating the same methods over a protein split, we find a
substantial performance difference in comparison to the splitting
scheme used in the original evaluation of each method. DTI-CDF was
originally evaluated on all three splitting schemes underlining this
point \cite{}. While \name provides comparable performance to the
na\"ive predictor and DTI-CDF in terms of MacroAUC, it yields
considerably better results with respect to $\textrm{MicroAUC}_p$. We
also find that methods that are trained using a protein split
generally result in higher $\textrm{MicroAUC}_p$ than methods trained
using a drug--target pair split.

@@

\todo[inline]{ add \name model to table to allow
  DIRECT comparison:}
The results of the analysis are summarized in the upper part of Figure
\ref{tab:Results}.

@@

\begin{itemize}
	\item \name comparison to other approaches on BioSnap prone to bias, as $ BioSnap: drugs 
	\times prots = 5500 \times 3000 $, while only 
	$\approx$ 1000 drugs have Sider Annotation, and 
\end{itemize}

>>>>

\documentclass{bioinfo}
\usepackage[english]{babel}
\copyrightyear{2015} \pubyear{2015}
\usepackage{natbib}
\bibliographystyle{natbib}

@@

\subtitle{Subject Section}
\newcommand{\name}{Voodoo}

\title[\name{}]{\name{}: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction}
\author[Sample \textit{et~al}.]{Tilman Hinnerichs\,$^{\text{\sfb 1,}*}$ and Robert Hoehndorf\,$^{\text{\sfb 2}}$}
\address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country and \\
$^{\text{\sf 2}}$Department, Institution, City, Post Code,
Country.}

@@

\section{Methods}
\subsection{Problem Description}
\name{} aims to solve the following problem: for a given drug and a
given protein we want to determine whether those interact or not.  We
do not differentiate between types of interaction such as activation
and inhibition, and do not predict the strength of the interaction.
If we additionally assume that our knowledge is complete and all
drug--protein pairs without a known interaction do not interact, we
can formulate the problem as a binary classification task.

@@

\subsection{\name{}: computational model to identify drugs that target a
  protein}

@@

We developed \name{} as a computational model to predict drug--target
interactions. Specifically, given a protein, \name{} will identify and
rank drugs that likely target this protein. \name{} combines two types
of features: structural information for drugs and proteins that can be
used to determine if the drug and protein physically interact, and
information about phenotypic effects of drugs and changes in protein
function that may ``localize'' on an interaction network (i.e.,
neighboring nodes will share some of these features or are
phenotypically similar).  As structural features, \name{} uses
structural representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name{} learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020} and
ontology-based annotations of drugs and proteins.

We construct a graph with proteins as nodes and protein-protein
interactions as edges, mapping the protein features to each target as
node features. \name{} then propagates information among the PPI network
utilizing graph convolutional steps, calculates the similarity of drug
and protein representations, and predicts whether there is an
interaction. The full workflow scheme is depicted in
figure~\ref{fig:ModelWorkflow}

@@

\begin{table*}[!tpb]
  \centering
	\begin{tabular}{|p{2.2cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|}
		\hline
		&\multicolumn{4}{c|}{(a) STITCH results}& \multicolumn{4}{c|}{(b) Yamanishi results}\\
		\hline
		\name{} results&\multicolumn{4}{c|}{PPI graph}&\multicolumn{4}{c|}{PPI graph}\\
		&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
		&Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		MolPred&$0.69$&$0.65$&$0.69$&$0.67$&$0.66$&$0.67$&$0.66$&$0.64$\\
		\hline
		OntoPred&$0.88$&$0.87$&$0.92$&$0.93$&$0.80$&$0.79$&$0.83$&$0.82$\\
		\hline
		\name{} (MolPred + OntoPred) & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$& $0.83$ & $0.82$&$\mathbf{0.84}$&$\mathbf{0.84}$\\
		\hline
	\end{tabular} \medskip

	\begin{tabular}{|l|p{1cm}|p{1cm}|p{1cm}|p{1cm}|}
		\hline
		(c) Approach&\multicolumn{2}{c|}{Original scheme}&\multicolumn{2}{c|}{Protein split}\\
		&Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		Naive predictor&DP pairs&$0.85$& --&--\\
		DTINet&DP pairs&$0.91$&$0.74$&$0.67$\\
		DTIGEMS+&DP pairs&$\mathbf{0.93}$& $0.72$& $0.68$ \\
		DTI-CDF&Proteins&$0.85$&$\mathbf{0.85}$&$0.79$\\
		\name{}&Proteins&$0.84$&$0.84$&$\mathbf{0.84}$\\
		\hline
		\hline
		(d) Approach&\multicolumn{2}{c|}{Original scheme}&\multicolumn{2}{c|}{Protein split}\\
		&Splitting scheme&Macro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		Naive predictor&DP pairs&$0.79$&-- &--\\
		DeepDTI&Drugs&$0.88$&$0.76$&$0.70$\\
		DeepDTA&DP pairs&$0.88$&$0.77$&$0.69$\\
		DeepConv-DTI&DP pairs&$0.88$&$0.76$&$0.73$\\
		MolTrans&DP pairs&$0.90$&$0.77$&$0.74$\\
		\hline
	\end{tabular}

	\caption{(a) + (b) Results for \name{} on STITCH and Yamanishi dataset evaluated with a 5-fold cross-validation. We hereby denote the molecular feature based predictor with \textit{MolPred}, while abbreviating the ontology based, top-down predictor with \textit{OntoPred}. (c) + (d) Results for various state of the art (c) drug--target interaction prediction methods on Yamanishi dataset and (d) drug--target affinity prediction methods on BIOSNAP dataset, evaluated on their original and the protein cross-validation splitting scheme, approximately reproducing the results of MolTrans \textbf{CITATION}.}
	\label{tab:Results}
\end{table*}

@@

\subsection{Protein-centric evaluation}
The goal of \name{} is to find candidate drugs that target a specific
protein; however, so far, we do not evaluate this application but
rather how \name{} would perform in finding plausible drug--target
interactions among all possible interactions (since we use the
MacroAUC as our main evaluation measure). This evaluation does not
correspond to the application of \name{} in finding drugs that target a
specific protein.  To provide a better estimate on how \name{} performs
for individual protein targets, we use micro-averages between proteins
and compute the \textit{MicroAUC} (see Section \ref{sec:evaluation});
to determine {\em MicroAUC}, we average the performance (true and
false positive rates) per protein instead of across all drug--protein
pairs; the resulting measure can therefore better estimate how \name{}
performs when tasked with finding a drug that targets a specific
protein.

@@

The way in which we formulated the na\"ive predictor, it is not
applicable for a protein split cross-validation as the number of
interactions for each protein in the validation set is unknown.  We
apply this naive predictor on both the STITCH and Yamanishi datasets,
using the full datasets as well as a 5-fold cross-validation over
drugs and over drug--protein pairs to compare the prediction results
directly to \name{}.  For each fold, we gradually increase $k$ to
determine the best performance for each fold. Using the full dataset,
drug--target split, and drug split, we obtain the following MacroAUC
results\todo[inline]{Put in Results table, then remove from here}: for
the STITCH database, we obtain a performance of $0.76$ on the whole
dataset, $0.70$ for the drug--target pairs and $0.73$ in case of the
drug splitting scheme; on the Yamanishi dataset, we obtain MacroAUC
scores of $0.88$, $0.84$ and $0.85$, for the total dataset,
drug--target pair and drug split, respectively.  The na\"ive predictor
shows higher performance on the Yamanishi dataset than on STITCH, and
a substantial gain in comparison to an expected random predictor on
both datasets. In the following, we utilize this nai\"ve predictor as
baseline to compare its performance to state of the art models and
\name{}.

@@

The results of our experiments are summarized in part (c) of Table
\ref{tab:Results}\todo{Check after restructure}; we calculated the
performance of all compared methods over their original splitting
scheme and over a protein split. We find that there is a large
difference in performance when evaluating over a drug--target pair
split compared to a protein split, with generally higher performance
achieved when using the drug--target pair split.  Second, when
evaluating the same methods over a protein split, we find a
substantial performance difference in comparison to the splitting
scheme used in the original evaluation of each method. DTI-CDF was
originally evaluated on all three splitting schemes underlining this
point \cite{}. While \name{} provides comparable performance to the
na\"ive predictor and DTI-CDF in terms of MacroAUC, it yields
considerably better results with respect to $\textrm{MicroAUC}_p$. We
also find that methods that are trained using a protein split
generally result in higher $\textrm{MicroAUC}_p$ than methods trained
using a drug--target pair split.

@@

\todo[inline]{ add \name{} model to table to allow
  DIRECT comparison:}
The results of the analysis are summarized in the upper part of Figure
\ref{tab:Results}.

@@

\begin{itemize}
	\item \name{} comparison to other approaches on BioSnap prone to bias, as $ BioSnap: drugs 
	\times prots = 5500 \times 3000 $, while only 
	$\approx$ 1000 drugs have Sider Annotation, and 
\end{itemize}
